Supplementary material for ``Screening Sinkhorn Algorithm for Regularized Optimal Transport''

### 2019-09-24 final_NeurIPS/appendix.tex
update appendix
add final NeurIPS

\subsection{Additional remarks}

It is worth to note that standard Sinkhorn's alternating minimization cannot be applied for the constrained screened dual problem~\eqref{screen-sinkhorn_second_def}. This appears more clearly while writing its optimality conditions (see Equations~\eqref{i-th-marginal-nu} and~\eqref{i-th-marginal-mu} ).
We resort to a L-BFGS-B algorithm to solve the constrained convex optimization problem on the screened variables~\eqref{screen-sinkhorn_second_def}, but any other efficient solver (e.g., proximal based method or Newton method) could be used. 
We further  emphasize that, as for the \textsc{Sinkhorn} algorithm, the \textsc{Screenkhorn} algorithm can be accelerated using a GPU implementation\footnote{\url{https://github.com/nepluno/lbfgsb-gpu}} of the L-BFGS-B algorithm~\cite{fei2014-GPU-L-BFGS-B}.

We have considered experiments with \textsc{Greenkhorn} algorithm~\cite{altschulernips17} but the implementation in POT library and our Python version of Matlab Alschuler's Greenkhorn code\footnote{\url{https://github.com/JasonAltschuler/OptimalTransportNIPS17}} were not competitive with \textsc{Sinkhorn}. Hence, for both versions, \textsc{Screenkhorn} is more competitive than \textsc{Greenkhorn}. The computation time gain reaches an order of $30$ when comparing our method with \textsc{Greenkhorn} while \textsc{Screenkhorn} is almost $2$ times faster than $\textsc{Sinkhorn}$.

@@

	All computations have been run on each single core of an Intel Xeon E5-2630 processor clocked at
	2.4 GHz in a Linux machine with 144 Gb of memory.
	The code that generates all figures given in the paper is available from \url{https://github.com/mzalaya/screenkhorn}.

}

>>>>

@@

	All computations have been run on each single core of an Intel Xeon E5-2630 processor clocked at
	2.4 GHz in a Linux machine with 144 Gb of memory.

It is worth to note that standard Sinkhorn's alternating minimization cannot be applied for the constrained screened dual problem~\eqref{screen-sinkhorn_second_def}. This appears more clearly while writing its optimality conditions (see Equations~\eqref{i-th-marginal-nu} and~\eqref{i-th-marginal-mu} ).
We resort to a L-BFGS-B algorithm to solve the constrained convex optimization problem on the screened variables~\eqref{screen-sinkhorn_second_def}, but any other efficient solver (e.g., proximal based method or Newton method) could be used. 
In our implementation, we use a tricky method, called \emph{restricted Sinkhorn} (see Algorithm~\ref{restricted_sinkhorn}) as a warm-start initialized point for the L-BFGS-B.
We further  emphasize that, as for the \textsc{Sinkhorn} algorithm, the \textsc{Screenkhorn} algorithm can be accelerated using a GPU implementation\footnote{\url{https://github.com/nepluno/lbfgsb-gpu}} of the L-BFGS-B algorithm~\cite{fei2014-GPU-L-BFGS-B}.

\LinesNotNumbered
\begin{algorithm}[tbp]
\SetNlSty{textbf}{}{.}
\DontPrintSemicolon
\caption{\textsc{Restricted \textsc{Sinkhorn}}$(K,\mu, \nu, u, v)$}
\label{restricted_sinkhorn}
\nl \textbf{set: } $\bar{f}_u = \varepsilon\kappa\, c(K_{I_{\varepsilon, \kappa}, J^\complement_{\varepsilon, \kappa}}), \bar{f}_v  = \varepsilon\kappa^{-1} \,r(K_{I^\complement_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}});$\\
% \nl $\bar{f}_u = \varepsilon\kappa\, c(K_{I_{\varepsilon, \kappa}, J^\complement_{\varepsilon, \kappa}});$\\
% \nl $\bar{f}_v  = \varepsilon\kappa^{-1} \,r(K_{I^\complement_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}});$\\
\nl \For{$t=1,2,3$}{
	$f^{(t)}_v \gets K^\top_{I_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}} u + \bar{f}_v;$\\
	$v^{(t)} \gets \frac{\nu_{J_{\varepsilon, \kappa}}}{\kappa f^{(t)}_v} ;$\\% / (\kappa K{IJ}_v);$\\
	$f^{(t)}_u \gets K_{I_{\varepsilon, \kappa}, J_{\varepsilon, \kappa}} v + \bar{f}_u;$\\
	$u^{(t)} \gets \frac{\kappa \mu_{I_{\varepsilon, \kappa}}}{f^{(t)}_u};$\\
	$u \gets u^{(t)}$, $v \gets v^{(t)}$;
}
\nl \Return{$(u^{(t)}, v^{(t)})$}
\end{algorithm}

We have considered experiments with \textsc{Greenkhorn} algorithm~\cite{altschulernips17} but the implementation in POT library and our Python version of Matlab Alschuler's Greenkhorn code\footnote{\url{https://github.com/JasonAltschuler/OptimalTransportNIPS17}} were not competitive with \textsc{Sinkhorn}. Hence, for both versions, \textsc{Screenkhorn} is more competitive than \textsc{Greenkhorn}. The computation time gain reaches an order of $30$ when comparing our method with \textsc{Greenkhorn} while \textsc{Screenkhorn} is almost $2$ times faster than $\textsc{Sinkhorn}$. 
}

\begin{figure*}[htbp]
	\centering
	\includegraphics[width=6.cm]{./figs/timegain_Screenkhorn_Greenkhorn.pdf}
	\caption{\ar{$\frac{T_{\textsc{Greenkhorn}}}{T_{\text{\textsc{Screenkhorn}}}}$: Running time gain  for the toy problem (see Section~\ref{subsec:analysing_toy_problem}) as a function of the data decimation factor in \textsc{Screenkhorn}, for different settings of the regularization parameter $\eta$.
		\label{fig:screenkhorn_greenkhorn_gain}}}
\end{figure*}

### 2019-09-24 final_NeurIPS/section_4_marginal_viols.tex
add final NeurIPS

Our second theoretical result, Proposition~\ref{prop:objective-error}, is an upper bound of the difference between objective values of \textsc{Screenkhorn} and dual of Sinkhorn divergence~\eqref{sinkhorn-dual}. 
\begin{proposition}
\label{prop:objective-error}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal pair solution of problem~\eqref{screen-sinkhorn_second_def} and $(u^\star, v^\star)$ is the pair solution of dual of Sinkhorn divergence~\eqref{sinkhorn-dual}. Then we have 
\begin{align*}
\Psi_{\varepsilon, \kappa}(u^{\text{sc}} ,v^{\text{sc}}) -\Psi(u^\star, v^\star)
= \bigO\Big(\Big(\frac{\norm{C}_\infty}{\eta} + \log\Big(\frac{(n\vee m)^2}{nmc_{\mu\nu}^{7/2}}\Big)\Big)(\norm{\mu - \mu^{\text{sc}}}_1 + \norm{\nu - \nu^{\text{sc}}}_1 + \omega_{\kappa}\Big).
\end{align*}
where $\omega_{\kappa} = (1- \kappa)\norm{\mu^{\text{sc}}}_1 + (\kappa^{-1} - 1)\norm{\nu^{\text{sc}}}_1 + \kappa^{-1} - \kappa$.
\end{proposition}
Proof of Proposition~\ref{prop:objective-error} is exposed in the supplementary material.
Comparing to some other analysis results of this quantity, see for instance Lemma 2 in~\cite{dvurechensky18aICML} and Lemma 3.1 in~\cite{lin2019}, our bound involves an additional term $\omega_{\kappa}$ (with $\omega_1 =0)$, that tends to zero as the pair budget $(n_b,m_b)$ goes to the full number budget of points $(n,m)$ (i.e., $\kappa$ goes to $1).$ 
\ar{To more characterize $\omega_\kappa$, a control of the $\ell_1$-norms of the screened marginals $\mu^{\text{sc}}$ and $\nu^{\text{sc}}$ are given in Equations~\eqref{l1-norm-mu-sc} and~\eqref{l1-norm-nu-sc} in Lemma~\ref{lemma_bounds_on_marginals} in the supplementary material.}

>>>>

Our second theoretical result, Proposition~\ref{prop:objective-error}, is an upper bound of the difference between objective values of \textsc{Screenkhorn} and dual of Sinkhorn divergence~\eqref{sinkhorn-dual}. 
\begin{proposition}
\label{prop:objective-error}
Let $(u^{\text{sc}}, v^{\text{sc}})$ be an optimal pair solution of problem~\eqref{screen-sinkhorn_second_def} and $(u^\star, v^\star)$ is the pair solution of dual of Sinkhorn divergence~\eqref{sinkhorn-dual}. Then we have 
\begin{align*}
\Psi_{\varepsilon, \kappa}(u^{\text{sc}} ,v^{\text{sc}}) -\Psi(u^\star, v^\star)
= \bigO\Big(\Big(\frac{\norm{C}_\infty}{\eta} + \log\Big(\frac{(n\vee m)^2}{nmc_{\mu\nu}^{7/2}}\Big)\Big)(\norm{\mu - \mu^{\text{sc}}}_1 + \norm{\nu - \nu^{\text{sc}}}_1 + \omega_{\kappa}\Big).
\end{align*}
where $\omega_{\kappa} = (1- \kappa)\norm{\mu^{\text{sc}}}_1 + (\kappa^{-1} - 1)\norm{\nu^{\text{sc}}}_1 + \kappa^{-1} - \kappa$.
\end{proposition}
Proof of Proposition~\ref{prop:objective-error} is exposed in the supplementary material.
Comparing to some other analysis results of this quantity, see for instance Lemma 2 in~\cite{dvurechensky18aICML} and Lemma 3.1 in~\cite{lin2019}, our bound involves an additional term $\omega_{\kappa}$ (with $\omega_1 =0)$, that tends to zero as the pair budget $(n_b,m_b)$ goes to the full number budget of points $(n,m)$ (i.e., $\kappa$ goes to $1).$ 
\ar{To more characterize $\omega_\kappa$, a control of the $\ell_1$-norms of the screened marginals $\mu^{\text{sc}}$ and $\nu^{\text{sc}}$ are given in Lemma~\ref{lemma_bounds_on_marginals} in the supplementary material.}
% Equations~\eqref{l1-norm-mu-sc} and~\eqref{l1-norm-nu-sc} in

### 2019-09-24 final_NeurIPS/section_5_numeric_expes.tex
add final NeurIPS

\subsection{Analysing on toy problem}

@@

	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\caption{OT Domain adaptation : running time gain  for MNIST as a function of the number of examples and the data decimation factor \textsc{Screenkhorn}. Hyperparameter values (left) $1$. (right) $10$.}
	\label{fig:otda}
\end{figure*}

>>>>

\subsection{Analysing on toy problem}
\label{subsec:analysing_toy_problem}

@@

	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	% --- \includegraphics[width=0.43\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl1.pdf}\hspace{1cm}
	\includegraphics[width=0.35\textwidth]{./figs/da_gain_mnist_regcl10.pdf}
	\caption{OT Domain adaptation : running time gain  for MNIST as a function of the number of examples and the data decimation factor in \textsc{Screenkhorn}. Hyperparameter values (left) $1$. (right) $10$.}
	\label{fig:otda}
\end{figure*}
